---
title: "Homework 1 - readable and efficient R code"
author: "Owen Jiang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1 - "function-alize" this code

Read over the code below and perform the following:

* Wrap it into a function `foobar0` which has arguments `x` and `z`
  and which returns the vector `x` at the end of the following code.
* Rewrite this into a function `foobar` which is easier to read, by
  reducing repetitive code. E.g. `foobar` might call a function to
  check the input, and another function to perform the three lines of
  computation. 
* Check that the two versions produce the same output using the
  function `all.equal`.

```{r}
## wrap it into a function foobar0
foobar0 <- function(x, z){
  if (sum(x >= .001) < 1) {
    stop("step 1 requires 1 observation(s) with value >= .001")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- sin(r) + .01
  if (sum(x >= .002) < 2) {
    stop("step 2 requires 2 observation(s) with value >= .002")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 2 * sin(r) + .02
  if (sum(x >= .003) < 3) {
    stop("step 3 requires 3 observation(s) with value >= .003")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 3 * sin(r) + .03
  if (sum(x >= .004) < 4) {
    stop("step 4 requires 4 observation(s) with value >= .004")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 4 * sin(r) + .04
  return(x)
}


## rewrite it into a function foobar
CheckInput <- function(x, n){
  threshold <- n/1000
  if (sum(x >= threshold) < n){
    stop(sprintf('step %s requires %s observation(s) with value >= %s', n, n, threshold))
  }
}


Computation <- function(x, z, n){
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- n * sin(r) + n/100
  return(x)
}


foobar <- function(x, z){
  n <- 1
  while(n <= 4){
    CheckInput(x, n)
    x <- Computation(x, z, n)
    n <- n + 1
  }
  return(x)
}


## check identical results
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
results.foobar0 <- foobar0(x, z)
results.foobar <- foobar(x, z)
all.equal(results.foobar0, results.foobar)
```

# Question 2 - vectorize this code and benchmark

* Take the following function `f0` and rewrite it as a function `f`,
  which is faster and easier to read, by removing the loop of `i` from
  1 to `m`.
* Benchmark `f` and `f0` using `microbenchmark`. How much faster is `f`?

```{r}
n <- 30
p <- 50
p2 <- 25
m <- 1000
set.seed(1)
x <- matrix(rnorm(n*p),nrow=n,ncol=p)
b <- matrix(rnorm(m*p),nrow=m,ncol=p)
a <- matrix(rnorm(m*p2),nrow=m,ncol=p2)
f0 <- function(x,b,a) {
  out <- numeric(0)
  for (i in seq_len(m)) {
    bb <- b[i,]
    aa <- a[i,]
    out <- c(out, sum(x %*% bb) + sum(aa))
  }
  out
}

## rewrite as a function f
f <- function(x, b, a){
  out <- colSums(tcrossprod(x, b)) + colSums(t(a))
  return(out)
}


## compare results and benchmark
result.f0 <- f0(x, b, a)
result.f <- f(x, b, a)
all.equal(result.f0, result.f)

library(microbenchmark)
microbenchmark(f0(x, b, a), f(x, b, a))
```
On average, `f` is 4 tiems faster than `f0`.


# Question 3 - build a faster t-test

* Rewrite the following function `getT0` which computes `m`
  two-sample t-tests (equal variance) between two groups as a function
  `getT`, which is faster by using vectorized operations over the `m`
  sets of observations. (There are functions in R packages, such as
  `genefilter::rowttests` which will quickly perform this operation,
  but I want you to write your own function using simple R functions
  like `rowSums`, etc.)
* Benchmark `getT` and `getT0`. How much faster is `getT`?

```{r}
m <- 400
n <- 50
little.n <- n/2
set.seed(1)
x <- matrix(rnorm(m*n),nrow=m,ncol=n)
f <- gl(2,little.n)
getT0 <- function(x, f) {
  ts <- sapply(seq_len(m), function(i) t.test(x[i,] ~ f, var.equal=TRUE)$statistic)
  unname(ts)
}


## rewrite a function getT
getT <- function(x, f){
  # assume f has two levels
  n <- ncol(x) 
  first.group <- which(f == levels(f)[1])
  X <- x[, first.group]
  Y <- x[, -first.group]
  X.mean <- rowMeans(X)
  Y.mean <- rowMeans(Y)
  sigma2.X <- rowSums((X - X.mean)^2) / (n / 2 - 1)
  sigma2.Y <- rowSums((Y - Y.mean)^2) / (n / 2 - 1)
  t.stat <- (X.mean - Y.mean) / (sqrt(2 / n * (sigma2.X + sigma2.Y)))
  return(t.stat)
}


## compare results and benchmark
result.getT0 <- getT0(x, f)
result.getT <- getT(x, f)
all.equal(result.getT0, result.getT)

microbenchmark(getT0(x, f), getT(x, f))

```
On aveage, `getT` is many orders of magnitude faster than `getT0`.